---
title: "Week 1 Module 1 - Introduction to R"

output: pdf_document
---

## For learners

Hi everyone! Welcome to the first class

Teaching coding can be challenging. So, here are some tips and built-in checkpoints that I have implemented to give everyone as smooth of an experience as possible.

* After every section (at least every other code chunk), I will stop and check in with you all and make sure your code is running properly. If I have not checked in, please raise your hand and I will remember to check in
* If, at any point, your code is giving you an error message, please attempt to continue with the code-along lecture. I will check in with you immediately after finishing my thought or after the active code chunk is run (max ~2 minutes; please be patient!).
  * If a lot of students have the same error, I will take more time to debug.
  * However, it is most likely a typo (e.g., wrong spelling, forgot a comma, forgot a closing bracket, etc.). Part of coding is recognizing the errors in your code! As you are waiting for me to come to you, please make sure that you are typing the code completely and error-free. And, chances are I will have answered your question or found a solution to the issue when I'm done explaining each small section!
  * If the current code is not necessary to run the code afterwards, I will verbalize that; I can help you debug after the lesson is over.
* After giving you time to complete the in-class exercises, I will check-in to see if you are done or if you need more time.

If any of these checkpoints are not working for you, please let me know and we can modify :)

## Introduction to R

Welcome to RStudio!

### Working with computers

Before we get into more practical matters, we want to provide a brief
background to the idea of working with computers. Essentially, computer
work is about humans communicating with a computer by modulating flows
of current in the hardware in order to get the computer to carry out
advanced calculations that we are unable to efficiently compute
ourselves. Early examples of human computer communication were quite
primitive and included physically disconnecting a wire and connecting it
again in a different spot. Luckily, we are not doing this anymore;
instead we have programs with Graphical User Interface with menus and
buttons that enable more efficient human to computer communication.

### Graphical user interfaces vs. text based user interfaces

An example of such a program that I think many of you are familiar with
is spreadsheet software such as Excel. 

Spreadsheet software is great for viewing and entering small data sets
and creating simple visualizations fast. However, it can be tricky to
design publication-ready figures, create automatic reproducible analysis
workflows, perform advanced calculations, and reliably clean data sets.
Even when using a spreadsheet program to record data, it is often
beneficial to have some some basic programming skills to facilitate the
analyses of those data.

Typing Commands instead of searching for them in menus is a
more efficient approach to communicate with the computer and a powerful
way of doing data analysis.

### In RStudio

RStudio includes the R console, but also many other convenient
functionalities, which makes it easier to get started and to work with
R. When you launch RStudio, you will see four panels. Starting at the
top left and going clockwise, these panels are:

-   The text editor panel. This is where we can write scripts, i.e.
    putting several commands of code together and saving them as a text
    document so that they are accessible for later and so that we can
    execute them all at once by running the script instead of typing
    them in one by one.
-   The environment panel, which shows us all the files and objects we
    currently loaded into R.
-   The files-plots-help panel. This panel shows the files in the
    current directory (the folder we are working out of), any plots we
    make later, and also documentation for various packages and
    functions. Here, the documentation is formatted in a way that is easier to      read
    and also provides links to the related sections.
-   The console is another space we can input code, only now the code is
    executed immediately and doesn't get saved at the end.

To change the appearance of your RStudio, navigate to `Tools > Global Options > Appearance`. You can change the the font and size, and the editor theme. The default is "Textmate", but if you like dark mode, I recommend "Tomorrow Night Bright". You can also change how your panels are organized. I like to have my Console and History below my Source, and that way I can see my working environment next to my code. That way, I know if an error I am getting is because I am missing an object or I renamed something oddly. Let's change that now. I playing around with the appearance if you prefer a different layout or colour scheme. Do makes you the most productive!

Another very useful thing with RStudio is that you have access to some
excellent cheat sheets in PDF format straight from the menu:
`Help -> Cheatsheets`!    

In the RStudio interface, we will be writing code in a format called the
_____________.

In RStudio, you will find two types of areas.

This current area is for ________ 

Some common keyboard shortcuts that may be useful:

* insert a new code cell (or press the green c with a + in the console toolbar)
* run only the line your cursor is on
* run all lines in the cell (or press the green play button)

I encourage you to take advantage of this space to take notes throughout
the lessons, feel free to add/edit any of the text in this document!
Markdown text will not affect how the code is run.

The second area is called ___________
These are my notes

Code chunks can be added using:
Mac: command + option + i
Windows: control + alt + i

```{r}
x <- 5 # code
x*2
```

x <- 5 this area is notes

The <- is used to assign value in objects. While a single equal
sign can work as well, it is not good coding practice for R and should
be saved for specifying parameters within functions (we'll get back to
this)

### Exercise
Try editing the code above: multiply the value of x by 2.



Within code cells, you can also add comments by #.


To comment out multiple lines, use Ctrl+Shift+C


### Saving data and generating reports

To save our notes, code, and graphs, all we have to do is to save the R
Notebook file, and then we can open it in RStudio next time again.
However, if we want someone else to look at this, we can't always just
send them the R Notebook file, because they might not have RStudio
installed. Another great feature of R Notebooks is that it is really
easy to export them to HTML, MS word, or PDF documents with figures and
professional typesetting. There are actually many academic papers that
are written entirely in this format and it is great for assignments and
reports, for instance, you might use this for some of your progress
updates to my supervisor. Since R Notebook files convert to HTML, it is
also easy to publish simple good-looking websites in it, so example,
[this one](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/), where
you can see the code chunks embedded nicely within the text.

Let's try to create a document in R.

First, let's set up the yaml-block. This is found at the top of your
document, and it is where you specify the title of your document, what
kind of output you want, and a few other things such as author list and
date.


## Writing Code

Let's write our first script!

```{r}
dozen<-12
```

### General syntax

Multiple values can be stored into an object using the function `c()`
for combine or concatenate

```{r}
goodjob<-c("high",5)
print(goodjob)
```

Functions act on objects and can have additional parameters within the
round brackets to specify how the command is carried out

```{r}

```

All functions have default parameters that you can access using the help
panel (same area as the "Files" and "Plots" panel) or using a `?` before
the function name

```{r}

```

### Exercise

Running the following lines of code will produce an error. Edit the code
in the code chunk to correct the error.

```{r}
times_three <- c(1, 3, 9, 27)
```

```{r}

```

## Operations of objects

Many researchers are motivated to move their workflows into programming
languages to scale up the analysis.

Values that are stored in objects can all be acted on all at once
together ______

```{r}

```



## Installing packages from CRAN

Let's revisit functions and packages.

Functions are Packages

R includes many pre-installed functions ______

Most general use packages are hosted on the CRAN network. Another
package repository you will encounter later specifically for biological
applications is called Bioconductor. To download and install a package
from CRAN, we will use another aptly named packaged called
`install.packages()`

```{r}

```

Once you have it installed, you can "open" them. You need to activate
the library every new session of R. ______

For this tidyverse package, you will likely see some warning messages -- ______


```{r}
library(tidyverse)
```

## Importing data

Rather than entering you data manually, you are more likely to read in
data from an external source such as an output file from a machine or
data stored in an Excel table. R is pretty flexible with the files that
it can accept, but there are differences to how it is read in.

The recommended format is a `______` file. 

Other common file formats are ______

In this module and the next, we will be working with a dataset measuring
categorical and integral characteristics of abalone gathered in
Australia. The following information about the data is considered the
metadata

**Source:**

Data comes from an original (non-machine-learning) study: Warwick J
Nash, Tracy L Sellers, Simon R Talbot, Andrew J Cawthorn and Wes B Ford
(1994) "The Population Biology of Abalone (*Haliotis* species) in
Tasmania. I. Blacklip Abalone (*H. rubra*) from the North Coast and
Islands of Bass Strait", Sea Fisheries Division, Technical Report No. 48
(ISSN 1034-3288) <https://archive.ics.uci.edu/ml/datasets/abalone>

**Data Set Information:**

Predicting the age of abalone from physical measurements. The age of
abalone is determined by cutting the shell through the cone, staining
it, and counting the number of rings through a microscope -- a boring
and time-consuming task. Other measurements, which are easier to obtain,
are used to predict the age. Further information, such as weather
patterns and location (hence food availability) may be required to solve
the problem.

**Attribute Information:**

Given is the attribute name, attribute type, the measurement unit and a
brief description. The number of rings is the value to predict: either
as a continuous value or as a classification problem.

*Name / Data Type / Measurement Unit / Description*

-   Sex / nominal / -- / M, F, and I (infant)
-   Length / continuous / mm / Longest shell measurement
-   Diameter / continuous / mm / perpendicular to length
-   Height / continuous / mm / with meat in shell
-   Whole weight / continuous / grams / whole abalone
-   Shucked weight / continuous / grams / weight of meat
-   Viscera weight / continuous / grams / gut weight (after bleeding)
-   Shell weight / continuous / grams / after being dried
-   Rings / integer / -- / +1.5 gives the age in years

As this is a csv file, we will be using the aptly named function
`_____` to import the file into R


```{r}

```


We'll do a quick test to make sure that everyone is set up and ready to
go. By the end of the next module, you'll be able to understand each piece of
the code and build it yourself!

```{r}
library(ggplot2)

ggplot(data = abalone,
       mapping = aes(x = Length, y = Whole.weight, col = Sex)) +
  geom_point(alpha = 0.3)
```



## Practice Problems


1.  The following code saves the average monthly temperature highs
in Toronto starting in January into an object. Identify the object,
function, and function parameter(s). What class will this object be?

```{r}
averageTorontoHigh <- c(0, 0, 5, 12, 19, 24, 27, 26, 23, 15, 9, 3)
```

> Answer: Object-> averageTorontoHigh
>         Class -> Numeric
>         Parameters -> (0, 0, 5, 12, 19, 24, 27, 26, 23, 15, 9, 3)



2.  Calculate the mean annual high temperature in Toronto. (Hint: what is the function to find the mean of an object?)

```{r}
mean(averageTorontoHigh)
```

3.  Windsor's monthly temperature highs are approximately 2 degrees
higher than Toronto. Create a new object called `averageWindsorHigh`
to store twelve values of monthly temperature high values based on
the data stored in the `averageTorontoHigh` object. Check your work
by printing out the values in the object

*Hint:* Remember you can do mathematical operations on objects.
*Hint:* Error messages are a normal process of coding and can be especially
frustrating when you're learning a new language. Don't worry! Every
coder (even those who have been coding for decades!) encounters error
messages when they are writing code. Keep at it!

```{r}
averageTorontoHigh <- c(0, 0, 5, 12, 19, 24, 27, 26, 23, 15, 9, 3)
averageWindsorHigh = averageTorontoHigh+2
print(averageWindsorHigh)
```

4.  When checking your work for the previous question, you may have
encountered error messages similar to the one listed below:

> Error: object " averagewindsorhigh\` not found

Describe two possible reasons that this error message was triggered.

>typo in object name 'averagewindsorhigh' instead of averageWindsorHigh
