---
title: "Assignment 1 - Introduction and Wrangling"
author: 'Nobody'
date: '23Sept2025'
output:
  pdf_document: default
  html_document: default
---

## Set-Up

The dataset can be accessed here: https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29

The phenotype attributions are as follows: 

1) ID number
2) Diagnosis (M = malignant, B = benign)
3-32)

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" - 1)

Here is some code to read in the dataset

> bCancer <- read.csv(here("3_csv", "breastCancer.csv"), header=T)


## Questions

1. When installing a package with the `install.packages()` function, the incoming package name needs to be enclosed in quotations. Why is this? [1]

```{r Cell 1}
# 1.
# .packages function argument takes string, if package was not enclosed with quotations 
# then package would appear as a variable with no assigned value/ error not found
```

2. Create a vector object with a numeric value, a character value, and a logical value. What data type is this vector and why? [1]

```{r Cell 2}
# 2.
myVector = c (1:3,"hello",TRUE)
summary(myVector) #vector type is 'character'
# Vector data type is 'character' because of the character value store in vector; this 
# changes all other data types to character as well.
```


3. Create a new object called `recent_fruits`.
a) Create 5 rows with the rownames as different fruits and 3 columns named "colour", "shape", and "taste". Describe the fruit's colour and shape in one word, rank the taste out of 10. Fill out the table! [0.5]
b) Calculate the mean taste rank from your previous table. [0.5]
c) c) Write a function called "middle_mean" that removes the largest and smallest taste rating, and then calculates the mean. Use the function on your dataset. *Hint*: you may need any of these functions: min(), max(), nrow(), length(). [0.75]

```{r Cell 3}
#3a
recent_fruits<- data.frame( colour <- c("Red","Green","Purple","Yellow","Yellow"), 
                           shape <- c("Spherical","Bell-shaped","Spherical","Elongated","Spheroid"),
                           taste <- c(5,6,8,9,2)
  
)
rownames(recent_fruits) <- c("Apple", "Pear", "Grape","Banana","Lemon")
print(recent_fruits)

#3b
print(mean(recent_fruits$taste))

#3c
middle_mean <- function(x){
  values <- x[-c(which(x==max(x)),which(x==min(x)))]
  return (mean(values))
}
print(middle_mean(recent_fruits$taste))
```


4. Use the bCancer dataset.
a) Subset the samples with a mean radius less than 20 into a new object called `large_tumour`. [0.5]
b) From the `large_tumour` object, subset for the columns describing the `id`, `diagnosis`, `radius_mean`, `texture_mean`, `smoothness_mean`, and `compactness_mean`. Reshape the data into a longer format so that there are only four columns: `id`, `diagnosis`, `variable`, and `value`. Save this into a new object called `large_tumour_long`. [0.5]
c) Group your data by diagnosis and variable, and calculate the mean, median, max, min, standard deviation, and count the number of observations. Save this into a table object called `large_tumour_summary`. *Hint*: first, figure out what type of data format you may need (`large_tumour_long` or `large_tumour`), and then check that your data columns are the correct class. [0.5]
d) Which patients have the 5 largest `radius_mean`, and what is their diagnosis? Which patients have the 5 smallest `radius_mean`, and what is their diagnosis? What does this trend potentially tell you about the data? *Hint*: there are multiple measurements for each patient, so use mean() to average these measurements. [0.75]
```{r Cell 4}
# CODE CELL 4
setwd("C:/Users/Delwar/Desktop/Humber/BINF5003- Data Mining, Modeling, and Biostatistcs/Labs/Assignment_1")
library(tidyverse)
breastCancer <- read.csv("breastCancer.csv", header = TRUE)

#4a)
large_tumour = breastCancer[breastCancer$radius_mean<20,]

#4b)
large_tumour_long <- large_tumour[,c("id", "diagnosis", "radius_mean", "texture_mean",
                                     "smoothness_mean", "compactness_mean")]
large_tumour_long <- large_tumour_long %>% 
  pivot_longer(cols = "radius_mean":"compactness_mean",
               names_to = "variable",
               values_to = "value")

#4c)
large_tumour_summary<-large_tumour_long %>%
  group_by(diagnosis, variable) %>% #"Group your data by diagnosis and variable"
  summarise(mean = mean(value),
            median = median(value),
            max = max(value),
            min = min(value),
            standard_deviation = sd(value),
            length = length(value)
            )
print(large_tumour_summary) #there are 8 observations in 'large_tumour_summary' 
                            #('M' was observed 357 times)
                            #('B' was observed 167 times)
#4d)
Largest <- breastCancer[,c("id", "diagnosis", "radius_mean")] %>% 
  slice_max(radius_mean,n=5)
Smallest <- breastCancer[,c("id", "diagnosis", "radius_mean")] %>% 
  slice_min(radius_mean,n=5)
print(Largest) #the tumors with largest mean radius are all "M" (Malignant) diagnosis
print(Smallest) #the tumors with smallest mean radius are all "B" (Benign) diagnosis
```

